---
title: "data prep: post-lab practice for point pattern analysis"
author: "Casey O'Hara"
format: 
  html:
    embed-resources: true
    code-fold: true
editor: visual
execute: 
  echo: true
  message: false
  warning: false
---

```{r setup}
library(tidyverse)
library(sf)
library(here)
```

# Prep cities shapefile as points

<https://data.cdc.gov/500-Cities-Places/500-Cities-City-Boundaries/n44h-hy2j/about_data>

This city boundary shapefile was extracted from Esri Data and Maps for ArcGIS 2014 - U.S. Populated Place Areas. This shapefile can be joined to 500 Cities city-level Data (GIS Friendly Format) in a geographic information system (GIS) to make city-level maps.

**Last Updated** July 27, 2023

|                     |                                                     |
|---------------------|-----------------------------------------------------|
| Contact Name        | 500 Cities Public Inquiries                         |
| Contact Email       | [places\@cdc.gov](mailto:places@cdc.gov)            |
| Publisher           | Centers for Disease Control and Prevention          |
| Public Access Level | Public Domain                                       |
| Bureau Code         | Centers for Disease Control and Prevention (009:20) |
| Program Code        | HHS:Program Management (009:020)                    |

```{r}
city_centroid_f <- here('data/us_cities/city_centroids.gpkg')
if(!file.exists(city_centroid_f)) {
  cities_poly <- read_sf(here('data/us_cities/500Cities_City_11082016',
                              'CityBoundaries.shp'))
  write_sf(cities_poly, here('data/us_cities/500Cities_City_11082016.gpkg'))
  # st_crs(cities_poly) ### epsg:3857
  cities_centroid <- cities_poly %>% st_centroid()
  write_sf(cities_centroid, city_centroid_f)
}

cities_centroid <- read_sf(city_centroid_f)
```

# Prep timezones as polygons

Also clip to just lower 48 states...

```{r}
tz_poly <- read_sf(here('data/us_cities/tz_us/tz_us.shp'))
# st_crs(tz_poly)

tzid_out <- c('America/Sitka',
              'America/Anchorage',
              'America/Nome',
              'America/Juneau',
              'America/Metlakatla',
              'America/Yakutat',
              'America/Adak',
              'Pacific/Honolulu',
              'America/Puerto_Rico',
              'America/St_Thomas')

tz_rename <- c('New_York' = 'Eastern',
               'Detroit'  = 'Eastern',
               'Los_Angeles' = 'Pacific',
               'Chicago' = 'Central',
               'Menominee' = 'Central',
               'Denver' = 'Mountain',
               'Boise' = 'Mountain',
               'Indiana' = 'Eastern',
               'North_Dakota' = 'Central',
               'Kentucky' = 'Eastern')
tz_3857 <- tz_poly %>%
  st_transform(st_crs(cities_centroid)) %>%
  filter(!TZID %in% tzid_out) %>%
  mutate(TZID = str_remove(TZID, 'America/') %>% str_remove('/.+')) %>%
  mutate(TZID = ifelse(TZID %in% names(tz_rename), tz_rename[TZID], TZID))

tz_vec <- tz_3857$TZID %>% unique()
tz_map_unioned <- lapply(tz_vec, FUN = function(tz) {
  ### tz <- 'Eastern'
  map_tmp <- tz_3857 %>%
    filter(TZID == tz) %>%
    st_union() %>%
    st_sf() %>%
    mutate(TZID = tz)
}) %>%
  bind_rows()
  
# plot(tz_map_unioned)

write_sf(tz_map_unioned, here('data/us_cities/tz_us.gpkg'))
```

# Prep cities as points with timezones bound, for quicker activity

```{r}
if(!'ZONE' %in% names(cities_centroid)) {
  cities_centroid <- cities_centroid %>%
    st_join(tz_map_unioned) %>%
    mutate(ZONE = TZID) %>%
    filter(!is.na(ZONE))
  write_sf(cities_centroid, here('data/us_cities/city_centroids.gpkg'))
}
```

# Plot

```{r}
ggplot() +
  geom_sf(data = tz_map_unioned, aes(fill = TZID)) +
  geom_sf(data = cities_centroid %>%
            filter(POP2010 >= 100000), aes(fill = ZONE), shape = 21) +
  scale_fill_viridis_d()
```
